Revealing transient strain in geodetic data with Gaussian process regression

### 2017-09-16 RevisedSubmission/hines_hetland_2017_gji.tex

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient crustal strain rates. We consider transient strain rates to be any deviation from secular strain rates, and our attention is limited to horizontal strain rates in this study. We denote transient crustal strain rates as $\dot\varepsilon(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x}$ are spatial coordinates in $\mathbb{R}^2$, and $t$ is time. We determine $\dot\varepsilon(p)$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p)$. We make a prior assumption that each component of $\vec{u}$ is a Gaussian process,
\begin{equation}\label{eq:TransientDeformation}
u_i(p) \sim \mathcal{N}\left(0,C_{u_i}\right),
\end{equation}
where $C_{u_i}(p,p')$ is a covariance function indicating how we expect $u_i(p)$ to covary with $u_i(p')$. The function $C_{u_i}$ must be positive definite, which means that the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$ is positive definite for any set of ordered pairs $\mitbf{P}$ \citep[sec. 2.5]{Cressie1992}. We assume that $C_{u_i}$ can be separated into spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}\left((\vec{x},t),(\vec{x}',t')\right) = X_i(\vec{x},\vec{x}')T_i(t,t').
\end{equation}  
As long as the functions $X_i$ and $T_i$ are positive definite, $C_{u_i}$ is guaranteed to also be positive definite and thus a valid covariance function \citep[sec. 4.2.4]{Rasmussen2006}. The appropriate choice for $X_i$ and $T_i$ may vary depending on the geophysical signal we are trying to describe, and we discuss this matter in Section \ref{sec:SignalModel}.

A posterior estimate of transient displacement is formed by conditioning $\vec{u}$ with GNSS observations. GNSS data records transient displacements as well as other physical and non-physical processes that we are not interested in. We describe GNSS observations at position $\vec{x}_i$ and time $t_j$ as a realization of the random variable 
\begin{align}\label{eq:Data}
\begin{split}
\vec{d}_{ij} = &\vec{u}(\vec{x}_i,t_j) + \vec{\eta}(\vec{x}_i,t_j) + \vec{w}_{ij} + \vec{a}^{(1)}_i + \vec{a}^{(2)}_it_j + \\
               &\vec{a}^{(3)}_i\sin(2 \pi t_j) + \vec{a}^{(4)}_i\cos(2 \pi t_j) + \vec{a}^{(5)}_i\sin(4 \pi t_j) + \vec{a}^{(6)}_i\cos(4 \pi t_j), 
\end{split}
\end{align}
where $\vec{a}^{(1)}_{i}$ is an offset that is unique for each GNSS station, $\vec{a}^{(2)}_{i}$ is the secular velocity at $\vec{x}_i$, and the sinusoids describe seasonal deformation (using units of years for $t_j$). We use $\vec{w}_{ij}$ to denote normally distributed, uncorrelated noise. Correlated noise which does not have a parametric representation is denoted by $\vec{\eta}$.  For example, $\vec{\eta}$ can consist of temporally correlated noise describing benchmark wobble \citep[e.g.,][]{Wyatt1982,Wyatt1989}, and/or spatially correlated noise describing common mode error \citep[e.g.,][]{Wdowinski1997}. For now, we will assume that each component of $\vec{\eta}$ is a Gaussian process distributed as $\eta_i \sim \mathcal{N}(0,C_{\eta_i})$, and $C_{\eta_i}$ is a spatially and temporally dependent covariance function. We consider the coefficients in eq. (\ref{eq:Data}) to be uncorrelated random variables distributed as $\mathcal{N}(0,\kappa^2)$ in the limit as $\kappa \to \infty$ (i.e., the coefficients have diffuse priors). Of course, the secular velocities, $\vec{a}^{(2)}_{i}$, are spatially correlated and we could invoke a tectonic model to form a prior on $\vec{a}^{(2)}_{i}$. However, in our application to the Pacific Northwest, we will be using displacement time series which are long enough to sufficiently constrain $\vec{a}^{(2)}_{i}$ for each station, avoiding the need to incorporate a prior. Likewise, seasonal deformation is spatially correlated \citep{Dong2002,Langbein2008}, and it may be worth exploring and exploiting such a correlation in a future study. We ignore any covariances between the easting and northing components of $\vec{d}_{ij}$. This simplifies our analysis by allowing us to constrain each component of transient displacements independently.

Each component of the posterior transient displacements, denoted as $\hat{u}_i$, is constrained by conditioning $u_i$ with GNSS observations of displacements along direction $i$, which we write as $\mitbf{d}^*_i$. The data vector consists of $n$ observations, which have been made at $m$ stations. We let $\mitbf{P}$ be the set of $(\vec{x}_i, t_j)$ pairs describing where and when each of the observations have been made. We also let $\mitbf{a}_i$ be the length $6m$ vector of coefficients from eq. (\ref{eq:Data}) for each station and for direction $i$. We use $\mitbf{G}$ to represent the $n \times 6m$ matrix of corresponding basis functions evaluated at each point in $\mitbf{P}$. We can then view $\mitbf{d}^*_i$ as a realization of the random vector
\begin{equation}
\mitbf{d}_i = u_i(\mitbf{P}) + \eta_i(\mitbf{P}) + \mitbf{G}\mitbf{a}_i,
\end{equation}
which is distributed as $\mathcal{N}(\mitbf{0},\mitbf{\Sigma} + \kappa^2\mitbf{G}\mitbf{G}^T)$, where
\begin{equation}\label{eq:Cd}
\mitbf{\Sigma} = C_u(\mitbf{P},\mitbf{P}) + C_\eta(\mitbf{P},\mitbf{P}) + 
              \mathrm{diag}\left(\mitbf{\sigma}^2\right).  
\end{equation}
It should be understood that notation such as $u(\mitbf{P})$ and $C_u(\mitbf{P},\mitbf{P})$ represents the column vector $[u(P_i)]_{P_i \in \mitbf{P}}$ and the matrix $[C_u(P_i,P_j)]_{(P_i,P_j) \in \mitbf{P} \times \mitbf{P}}$, respectively.

>>>>

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient strain rates. We consider transient strain rates to be any deviation from secular strain rates, and our attention is limited to horizontal strain rates. We denote transient strain rates as $\dot\varepsilon(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x}$ are spatial coordinates in a 2-D map projection aligned with the cardinal directions (i.e, $\vec{x} = (x_\mathrm{e},x_\mathrm{n})$), and $t$ is time. We determine $\dot\varepsilon$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p)$. Transient displacements have an easting and northing component (i.e., $\vec{u} = (u_\mathrm{e}, u_\mathrm{n})$), and we make a prior assumption that each component of $\vec{u}$ is a Gaussian process,
\begin{equation}\label{eq:TransientDeformation}
u_i \sim \mathcal{N}\left(0,C_{u_i}\right),
\end{equation}
where $C_{u_i}(p,p')$ is a covariance function indicating how we expect $u_i(p)$ to covary with $u_i(p')$. The function $C_{u_i}$ must be positive definite, which means that the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$ is positive definite for any set of ordered pairs $\mitbf{P}$ \citep[sec. 2.5]{Cressie1992}. We assume that $C_{u_i}$ can be separated into spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}\left((\vec{x},t),(\vec{x}',t')\right) = X_i(\vec{x},\vec{x}')T_i(t,t').
\end{equation}  
As long as the functions $X_i$ and $T_i$ are positive definite, $C_{u_i}$ is guaranteed to also be positive definite and thus a valid covariance function \citep[sec. 4.2.4]{Rasmussen2006}. The appropriate choice for $X_i$ and $T_i$ may vary depending on the geophysical signal we are trying to describe, and we discuss this matter in Section \ref{sec:SignalModel}.

GNSS data records transient displacements as well as other physical and non-physical processes that we are not interested in. We describe GNSS displacement observations at position $\vec{x}_j$ and time $t_k$ as a realization of the random variable 
\begin{align}\label{eq:Data}
\begin{split}
\vec{d}_{jk} = &\vec{u}(\vec{x}_j,t_k) + \vec{\eta}(\vec{x}_j,t_k) + \vec{a}^{(1)}_j + \vec{a}^{(2)}_jt_k + \vec{a}^{(3)}_j\sin(2 \pi t_k) + \\ 
               &\vec{a}^{(4)}_j\cos(2 \pi t_k) + \vec{a}^{(5)}_j\sin(4 \pi t_k) + \vec{a}^{(6)}_j\cos(4 \pi t_k), 
\end{split}
\end{align}
where $\vec{a}^{(1)}_j$ is an offset that is unique for each GNSS station, $\vec{a}^{(2)}_j$ is the secular velocity at $\vec{x}_j$, and the sinusoids describe seasonal deformation (using units of years for $t_k$). We use $\vec{\eta}$ to represent all the stochastic noise processes. For example, $\vec{\eta}$ can consist of uncorrelated white noise, temporally correlated noise describing benchmark wobble \citep[e.g.,][]{Wyatt1982,Wyatt1989}, and/or spatially correlated noise describing common mode error \citep[e.g.,][]{Wdowinski1997}. For now, we only make the assumption that each component of $\vec{\eta}$ is a Gaussian process distributed as $\eta_i \sim \mathcal{N}(0,C_{\eta_i})$, and $C_{\eta_i}$ is a spatially and temporally dependent covariance function. We consider the coefficients in eq. (\ref{eq:Data}) to be uncorrelated random variables distributed as $\mathcal{N}(0,\kappa^2)$ in the limit as $\kappa \to \infty$ (i.e., the coefficients have diffuse priors). Of course, the secular velocities, $\vec{a}^{(2)}_j$, are spatially correlated and we could invoke a tectonic model to form a prior on $\vec{a}^{(2)}_j$. However, in our application to the Pacific Northwest, we will be using displacement time series which are long enough to sufficiently constrain $\vec{a}^{(2)}_j$ for each station, avoiding the need to incorporate a prior. Likewise, seasonal deformation is spatially correlated \citep{Dong2002,Langbein2008}, and it may be worth exploring and exploiting such a correlation in a future study.

To reduce the computational complexity of our analysis, we assume the easting and northing components of $\vec{u}$ are uncorrelated. Likewise, we ignore any covariances between the components in $\vec{\eta}$. By ignoring covariances between the easting and northing components, we are inherently assuming that the principle components of $\vec{u}$ and $\vec{\eta}$ are aligned with the cardinal directions, which is admittedly an arbitrary assumption.

We combine the prior on transient displacements with GNSS displacement data to form a posterior estimate of transient displacements. Each component of the posterior, denoted as $\hat{u}_i$, is constrained by conditioning $u_i$ with GNSS data for the corresponding direction. We denote the data vector consisting of $n$ GNSS displacement observations as $\mitbf{d}^*_i$. We let $\mitbf{P}$ be the set of $(\vec{x}_j, t_k)$ pairs describing where and when each of the observations have been made. We collect the coefficients from eq. (\ref{eq:Data}) into the vector $\mitbf{a}_i$, which contains 6 elements for each GNSS station. We use $\mitbf{G}$ to represent the matrix of corresponding basis functions evaluated at each point in $\mitbf{P}$. We can then view $\mitbf{d}^*_i$ as a realization of the random vector
\begin{equation}
\mitbf{d}_i = u_i(\mitbf{P}) + \eta_i(\mitbf{P}) + \mitbf{G}\mitbf{a}_i,
\end{equation}
which is distributed as $\mathcal{N}(\mitbf{0},\mitbf{\Sigma} + \kappa^2\mitbf{G}\mitbf{G}^T)$, where
\begin{equation}\label{eq:Cd}
\mitbf{\Sigma} = C_u(\mitbf{P},\mitbf{P}) + C_\eta(\mitbf{P},\mitbf{P}) + 
              \mathrm{diag}\left(\mitbf{\sigma}^2\right).  
\end{equation}
It should be understood that notation such as $u(\mitbf{P})$ and $C_u(\mitbf{P},\mitbf{P})$ represents the column vector $[u(P_i)]_{P_i \in \mitbf{P}}$ and the matrix $[C_u(P_i,P_j)]_{(P_i,P_j) \in \mitbf{P} \times \mitbf{P}}$, respectively.
